---
title: "Nematode Addition"
author: "Aaron Ferrucci"
date: "8/9/2020"
output:
  html_document:
    df_print: paged
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r message=F, warning=F, echo=F}
library(DiagrammeR)
```

From [538 Riddler: Can You Reach The Beach?](https://fivethirtyeight.com/features/can-you-reach-the-beach/)

> This week’s Riddler Classic is inspired by Kareem Carr:
> 
> We usually think of addition as an operation applied to a field like
> the rational numbers or the real numbers. And there is good reason for
> that — as Kareem says, "Mathematicians have done all the hard work
> of figuring out how to make calculations track with reality. They kept
> modifying and refining the number system until everything worked out. It
> took centuries of brilliant minds to do this!"
> 
> Now suppose we defined addition another (admittedly less useful)
> way, using a classic model organism: the nematode. To compute the sum
> of x and y, you combine groups of x and y nematodes and leave them for
> 24 hours. When you come back, you count up how many you have — and
> that’s the sum!
> 
> It turns out that, over the course of 24 hours, the nematodes pair up,
> and each pair has one offspring 50 percent of the time. (If you have an
> odd number of nematodes, they will still pair up, but one will be left
> out.) So if you want to compute 1+1, half the time you’ll get 2 and
> half the time you’ll get 3. If you compute 2+2, 25 percent of the time
> you get 4, 50 percent of the time you’ll get 5, and 25 percent of the
> time you’ll get 6.
> 
> While we’re at it, let’s define exponentiation for sums of
> nematodes. Raising a sum to a power means leaving that sum of nematodes
> for the number of days specified by the exponent.
> 
> With this number system, what is the expected value of (1+1)^4?
> 
> Extra credit: As N gets larger and larger, what does the expected value 
> of (1+1)^N approach?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basics: Nematode Addition
Nematodes pair up, and each pair produces a sub-sum of 2 or 3 with equal likelihood. A tree structure is a handy way to visualize the process. The parent node contains the standard sum, and children show the nematode sum outputs. Here's ``1 + 1``:
```{r echo = F, warning=FALSE}
a_graph <- create_graph() %>%
  add_node(label="1+1") %>%
  add_node(label="2") %>%
  add_node(label="3") %>%
  add_edge(from=1, to=2) %>%
  add_edge(from=1, to=3)
render_graph(a_graph, layout="tree")
```

A node with standard sum ``n`` will have $2^{\lfloor n/2 \rfloor}$ outputs of equal
likelihood. Odd sums ``m`` have the same number of outputs as ``m - 1``, with each output being one greater than that of ``m - 1``. The specific addends don't make any difference (that is, ``1 + 4`` produces the same outputs as ``2 + 3``) so from now on I'll put the standard sum in the parent node, rather than the addition expression. Some examples:

```{r echo = F, warning=FALSE}
a_graph <- create_graph() %>%
  add_node(label="3") %>%
  add_node(label="3") %>%
  add_node(label="4") %>%
  add_edge(from=1, to=2) %>%
  add_edge(from=1, to=3)
render_graph(a_graph, layout="tree")
```

```{r echo = F, warning=FALSE}
a_graph <- create_graph() %>%
  add_node(label="4") %>%
  add_node(label="4") %>%
  add_node(label="5") %>%
  add_node(label="5") %>%
  add_node(label="6") %>%
  add_edge(from=1, to=2) %>%
  add_edge(from=1, to=3) %>%
  add_edge(from=1, to=4) %>%
  add_edge(from=1, to=5)
render_graph(a_graph, layout="tree")
```

```{r echo = F, warning=FALSE}
a_graph <- create_graph() %>%
  add_node(label="5") %>%
  add_node(label="5") %>%
  add_node(label="6") %>%
  add_node(label="6") %>%
  add_node(label="7") %>%
  add_edge(from=1, to=2) %>%
  add_edge(from=1, to=3) %>%
  add_edge(from=1, to=4) %>%
  add_edge(from=1, to=5)
render_graph(a_graph, layout="tree")
```

## Exponentiation of Nematode Sums
Let's think about exponentiation. Each successive power adds a layer to the tree; the tree for a sum to the power ``n`` will have ``n + 1`` levels. Here's ``(1+1)^2``:
```{r echo = F, warning=FALSE}
a_graph <- create_graph() %>%
  add_node(label="2") %>%
  add_node(label="2") %>%
  add_node(label="3") %>%
  add_node(label="2") %>%
  add_node(label="3") %>%
  add_node(label="3") %>%
  add_node(label="4") %>%
  add_edge(from=1, to=2) %>%
  add_edge(from=1, to=3) %>%
  add_edge(from=2, to=4) %>%
  add_edge(from=2, to=5) %>%
  add_edge(from=3, to=6) %>%
  add_edge(from=3, to=7)
render_graph(a_graph, layout="tree")
```
So far we have a nice binary tree, but as soon as a node reaches value 4, the branching factor goes above 2. Here's ``(1 + 1)^3``:

```{r echo = F, warning=FALSE}
a_graph <- a_graph %>%
  add_node(label="2") %>%
  add_node(label="3") %>%
  add_node(label="3") %>%
  add_node(label="4") %>%
  add_node(label="3") %>%
  add_node(label="4") %>%
  add_node(label="4") %>%
  add_node(label="5") %>%
  add_node(label="5") %>%
  add_node(label="6") %>%
  add_edge(from=4, to=8) %>%
  add_edge(from=4, to=9) %>%
  add_edge(from=5, to=10) %>%
  add_edge(from=5, to=11) %>%
  add_edge(from=6, to=12) %>%
  add_edge(from=6, to=13) %>%
  add_edge(from=7, to=14) %>%
  add_edge(from=7, to=15) %>%
  add_edge(from=7, to=16) %>%
  add_edge(from=7, to=17)
  
render_graph(a_graph, layout="tree")
```

One more power to go! Here's ``(1 + 1)^4`` (I had to drop the 'tree' format, it renders poorly):

```{r echo = F, warning=T}
node0 <- c(2)
node1 <- c(node0, 3)
node2 <- c(node1, 3, 4)
node3 <- c(node2, 3, 4, 4, 5, 5, 6)
node4 <- c(node3, 3, 4, 4, 5, 5, 6, 4, 5, 5, 6, 5, 6, 6, 7, 5, 6, 6, 7, 6, 7, 7, 7, 8, 8, 8, 9)
from1 <- c(1, 1)
to1 <- seq(2, 3)
from2 <- c(2, 2, 3, 3)
to2 <- seq(4, 7)
from3 <- c(4, 4, 5, 5, 6, 6, 7, 7, 7, 7)
to3 <- seq(8, 17)
from4 <- c(8, 8, 9, 9, 10, 10, 11, 11, 11, 11, 12, 12, 13, 13, 13, 13, 14, 14, 14, 14, 15, 15, 15, 15, 16, 16, 16, 16, 17, 17, 17, 17, 17, 17, 17, 17)
to4 <- seq(18, 53)

nodelabels <- c(node0, node1, node2, node3, node4)
froms <- c(from1, from2, from3, from4)
tos <- c(to1, to2, to3, to4)

# ndf <- create_node_df(n=length(nodelabels), width=0.5, label=nodelabels, fontsize=50)
ndf <- create_node_df(n=length(nodelabels), width=1, label=nodelabels, fontsize=40)
edf <- create_edge_df(from=froms, to=tos)
a_graph <- create_graph(nodes_df=ndf, edges_df=edf)
render_graph(a_graph, width=1024, height=1024)
```

